Fast, Robust Classification with Applications to Periodic Variable Stars

### 2012-05-10 report/ms.tex
final change to ms

Periodic variables may be divided into a few dozen classes based on physical properties of light sources. Separating the sources into classes is a critical step in turning raw astronomical observations into scientific knowledge. The size of modern data sets requires that much of this work be automated by machine learning and statistical classifiers. 
\begin{figure}[ht]
\begin{minipage}[t]{.45\linewidth}
  \centering
    \begin{includegraphics}[scale=.48]{2000.pdf}
      \caption{Light curve of a Classical Cepheid type variable star. The brightness of the star is on the y-axis. For the top plot the brightness is plotted against time. This time series is periodic. Using fourier methods we can estimate a period of 4.51 days. We can convert the times from the top plot into phase ((time modulo period) / period)). Brightness versus phase is plotted in the bottom plot. Here we can observe the structure in the time series, which is usually similar for stars of the same class but different for stars of different classes.\label{fig:cepheid}}
    \end{includegraphics}
\end{minipage}
\hspace{1cm}
\begin{minipage}[t]{.45\linewidth}
  \centering
    \begin{includegraphics}[scale=.48]{204.pdf}
      \caption{Light curve of a Mira type variable star. Mira variable stars have higher amplitude and longer periods than Cepheid variables. Notice that the period estimate of 322.55 appears to be incorrect. The true period is likely half of this.\label{fig:mira}}
    \end{includegraphics}
\end{minipage}
\end{figure}

Figure \ref{fig:cepheid} displays the light curve (i.e. time series) of a periodic variable belonging to the class Classical Cepheid. The points in the top plot represent flux measurements in magnitudes (i.e. brightness of the source) made by the telescope at particular times. The 0 point on the time axis is arbitrary. Using fourier methods, one can estimate a period using these measurements.  The lower plot of Figure \ref{fig:cepheid} displays the flux measurements of the same object. However here the x-axis is phase of each time measurement, computed using the estimated period of 4.51 days. Here we can observe the structure of the periodic variation. This is known as the \textit{folded light curve}.

Figure \ref{fig:mira} displays an example of a Mira light curve. From the y-axis we can see that this source has higher amplitude than the Classical Cepheid (this is typical of the Mira class) and more sinusoidal variation (also typical). Note that the fourier methods appear to have estimated an incorrect period for this source. The true period appears to be around 161 days, half of the estimate.
\subsection{Classification Methodology}
There has been a lot of recent progress in the astronomy literature towards developing highly accurate classifiers for the time series generated by periodic variables \cite{debosscher2007automated,richards2011machine,dubath2011random}. The standard approach works as follows. A telescope observes a source $j$ at times $t_{1},\ldots,t_{l}$, recording flux measurements (how bright the object is) of $m_{1},\ldots,m_{l}$. Typically there are measurements of uncertainty on the flux measurements $e_{1},\ldots,e_{l}$. So each source $j$ is initially characterized by an $l$ tuples $D_j=\{(t_{i},m_{i},e_{i})\}_{i=1}^{l}$. Note that $l$, the number of times the source is observed, is different for each $D_j$. Also the time sampling of the flux measurements is irregular and different for each source. Associated with each source is a classification, such as Mira, Classical Cepheid, RR Lyrae, etc.

In order to construct a classifier, features are \textit{extracted} from $D_j$ (i.e. take functions of $D_j$) that will separate sources into different classes. Features vary from study to study, but typical ones include period (inverse of strongest fourier frequency), amplitude, skew, and estimates of derivatives.

@@

\subsection{Uncertain Features}
\begin{figure}[h]
  \begin{center}
    \begin{includegraphics}[scale=.5]{period_amplitude.pdf}
      \caption{Period-Amplitude relationship for two classes: \textit{Classical Cepheid} stars and \textit{Population II Cepheid} stars. The green box represents (roughly) the physically possible range of periods and amplitudes. Clearly some of these features have been estimated incorrectly.\label{fig:period_amplitude}}
    \end{includegraphics}
  \end{center}
\end{figure}
A major problem with the standard approach is the high levels and heteroskedastic nature of the uncertainty in the features due to having poorly sampled time series ($l$ is small), high error in the flux measurements ($e_{i}$ are systematically large), or irregular nature of sampling ($t_i$'s are highly concentrated, giving a poor representation of variation). This leads to situations where features meant to represent physical quantities of the time series, such as amplitude or period, are incorrect. As an example in Figure \ref{fig:period_amplitude} we plot the period and amplitude features for two classes of stars with a box around known physical limits for these features for these two classes. Some of the observations have values outside these physical limits, clearly indicating that the features are incorrectly estimated. This suggests that classifiers which incorporate feature uncertainty may be able to acheive improved performance.

@@

This model, but without the $\frac{\lambda}{2}\beta^T\beta$ regularization, was considered in El Ghaoui \cite{el2003robust}. It is easy to show that for this form of $\mathcal{X}(\rho)$ one can explicity solve for the inner max in equation \eqref{eq:svm_minimax} and obtain
\begin{equation}
\label{eq:svmInit}
\min_{\beta}\sum_{i=1}^n(1-y_i(\beta_0+\beta^Tx_i)+\rho\sigma_i^T|\beta|)_++\frac{\lambda}{2} ||\beta||_2^2.
\end{equation}
where $\sigma_i$ is the $i^{th}$ row of $\Sigma$. In \cite{el2003robust}, El Ghaoui drew similarities between the $\rho\sigma_i^T|\beta|$ in this expression and standard $L_1$ regularization for SVMs. For example, if $\Sigma_{i,j} = 1 \ \forall i,j$ then then expression becomes $\rho||\beta||_1$. This interpretation of $\mathcal{X}(\rho)$ in part motivates the use of $\rho$ as a tuning parameter rather than a quantity that is fixed by the inherent uncertainty in the data.

@@

From now on, we can safely assume that $n_+>n_-$, in which case the initial state is $\beta = 0$, $\beta_0=1$, $\xi_i = 0$ for $i\in \mathcal I_+$ and $\xi_i = 2$ for $i\in \mathcal I_-$. Thus by optimality conditions, for any $i\in \mathcal I_-$, $\gamma_i =0$ and so $\alpha_i=1$. In addition, since $0=\sum_{i=1}^n \alpha_i y_i = \sum_{i\in \mathcal I_+} \alpha_i - \sum_{i\in \mathcal I_-} \alpha_i$, we have $\sum_{i\in \mathcal I_+} \alpha_i = \sum_{i\in \mathcal I_-} \alpha_i = n_-$. The following lemma determines the initial value for the dual valuables:
\begin{lamma} Let $\tilde{\beta} = \sum_{i=1}^n \alpha_i y_i x_i + \nu - \mu$, 
\[
\begin{array}{rl}
(\alpha^*,\mu^*,\nu^*) = & \arg\min ||\tilde{\beta}||_2^2, \\
& \text{subject to: } 0\leq \alpha_i\leq 1,  \forall i \in \mathcal I_+; \alpha_i = 1,  \forall i \in \mathcal I_-; \\
& \sum_{i \in \mathcal I_+}\alpha_i = n_-, \rho\Sigma\alpha \geq \mu+\nu, \mu\geq0, \nu\geq0. \\
\end{array}
\]
 and $c^* = \rho\Sigma\alpha^* - \mu^*-\nu^*$. Then for some $\lambda_0$ we have for all $\lambda>\lambda_0$, $(\alpha(\lambda),\mu(\lambda),\nu(\lambda),c(\lambda)) = (\alpha^*, \mu^*,\nu^*,c^*)$.
\end{lamma}  
\noindent\textbf{Proof.} The proof follows directly from Lemma 2 in \cite{hastie2004entire}. The key observasion is that  $\sum_{i=1}^n \alpha_i = 2n_-$ and this sum will remain constant for a while as $\beta$ grows from zero. Then we can plug this sum into the dual problem \eqref{eq:dual} and solve it with additional constraints: $\alpha_i = 1,  \forall i \in \mathcal I_-$ and $\sum_{i \in \mathcal I_+}\alpha_i = n_-$. \\

@@

%Remarks: in \cite{hastie2004entire}, Hastie et.al. split the initialization step into two cases depending on whether $n_+=n_-$ or $n_+>n_-$. In their case $n_+=n_-$, the initial state can be obtained efficiently without solving the quadratic programming problem. This does not seem to be the case in our problem, as it also involves finding the initial values of the extra dual variables $\mu$ and $\nu$. Therefore solving a quardratic problem for starting up the path algorithm cannot be avoided in our case.\\

@@

%\subsection{Extension to Kernal} (You need to transform the uncertainty in the original data space into the kernal space)
%\note{in fitting the CEP data,  between 2.03 and 2.13 $\beta0$ value can be arbitrary}
\subsection{Matlab and R programs}
All the programs for this project are included in the {\tt zip} file submitted together with this hard copy. In particular, the matlab package {\tt svmIntervalPath} is the implementation of our path algorithm. Also, matlab and R code for testing and data analysis can be found in that file as well.

\section{Application to Variable Star Data Sets}
\label{sec:application}
We now return to the periodic variable classification problem and discuss construction of hyper-rectangles, the specific data sets we use, and results from the models.
\subsection{Constructing the Intervals from the Time Series}
\begin{figure*}[ht]
\begin{center}
{\small
\framebox[6in]{
\begin{minipage}[t]{5.8in}
% Insert here your text
\begin{center} \textbf{Constructing Hyper-rectangles from Time Series} \end{center}
\begin{enumerate}
\item Order the light curve measurements in time i.e. $\{(t_1,m_1,e_1),(t_2,m_2,e_2), \ldots , (t_l,m_l,e_l)\}$ where $t_i < t_j$ for $i < j$.
\item Slice each light curve into 5 contiguous sections, producing 5 time series. Slice 1 is $(t_i,m_i,e_i)$ for $0 \leq i \leq .5l$, slice 2 for $.125l \leq i \leq .625l$, slice 3 for $.25l \leq i \leq .75l$ ect.
%% \begin{enumerate}
%% \item $\{(t_1,m_1,e_1), \ldots , (t_{l/2},m_{l/2},e_{l/2})\}$
%% \item $\{(t_1,m_1,e_1),(t_2,m_2,e_2), \ldots , (t_l,m_l,e_l)\}$
%% \end{enumerate}
\item Derive features for each of these slices.
\item Put the interval minimum at the lowest value of the feature for the 5 slices. Put the interval max at the maximum feature value across the 5 slices.
\end{enumerate}
\end{minipage}
}
}
\end{center}
\caption{Description of interval construction algorithm.\label{fig:interval_construction}}
\label{lb}
\end{figure*}
In certain applications, methods for constructing hyper-rectangles (or hyper-spheres) in feature space may be fairly straitforward. For example in El Ghaoui \cite{el2003robust}, the authors discuss applications to micro-array data where several replicates of some experiment are available, and a hyper-rectangle can be chosen to enclose all replicates.

@@

\begin{figure}
\include{RR2_features}
\caption{RR features with non-zero coefficients. Some coefficient are reported here as 0 due to rounding. Choosing a slightly higher $\rho$ may have caused severl of the features with very small coefficients to become exactly 0, leading to a very sparse solution.\label{fig:RRfeatures}}
\end{figure}

\begin{figure}
\include{CEP_features}
\caption{CEP features. Features with large coefficients correspond well to features that are known to be important for separating these two classes.\label{fig:CEPfeatures}}
\end{figure}

@@

\subsection{Conclusions}
We have developed an algorithm and empirically studied an SVM interval model which incorporates feature uncertainty and produces sparse classifiers. The model improves on classification performance over ignoring feature uncertainty. We feel that the results warrent applying this method to larger and multi-class astronomy data sets. From a data perspective we faced several challenges, the greatest of which was determinining intervals for features. This will be common in many high dimensional classification problems where there is uncertainty on features, but no clear way to contruct intervals.

>>>>

Periodic variables may be divided into a few dozen classes based on physical properties of light sources. Separating the sources into classes is a critical step in turning raw astronomical observations into scientific knowledge. The size of modern data sets requires that much of this work be automated by machine learning and statistical classifiers. 
\begin{figure}[ht]
\begin{minipage}[t]{.45\linewidth}
  \centering
    \begin{includegraphics}[scale=.48]{2000.pdf}
      \caption{Light curve of a Classical Cepheid type variable star. The brightness of the star is on the y-axis. For the top plot the brightness is plotted against time. This time series is periodic. Using Fourier methods we can estimate a period of 4.51 days. We can convert the times from the top plot into phase ((time modulo period) / period)). Brightness versus phase is plotted in the bottom plot. Here we can observe the structure in the time series, which is usually similar for stars of the same class but different for stars of different classes.\label{fig:cepheid}}
    \end{includegraphics}
\end{minipage}
\hspace{1cm}
\begin{minipage}[t]{.45\linewidth}
  \centering
    \begin{includegraphics}[scale=.48]{204.pdf}
      \caption{Light curve of a Mira type variable star. Mira variable stars have higher amplitude and longer periods than Cepheid variables. Notice that the period estimate of 322.55 appears to be incorrect. The true period is likely half of this.\label{fig:mira}}
    \end{includegraphics}
\end{minipage}
\end{figure}

Figure \ref{fig:cepheid} displays the light curve (i.e. time series) of a periodic variable belonging to the class Classical Cepheid. The points in the top plot represent flux measurements in magnitudes (i.e. brightness of the source) made by the telescope at particular times. The 0 point on the time axis is arbitrary. Using Fourier methods, one can estimate a period using these measurements.  The lower plot of Figure \ref{fig:cepheid} displays the flux measurements of the same object. However here the x-axis is phase of each time measurement, computed using the estimated period of 4.51 days. Here we can observe the structure of the periodic variation. This is known as the \textit{folded light curve}.

Figure \ref{fig:mira} displays an example of a Mira light curve. From the y-axis we can see that this source has higher amplitude than the Classical Cepheid (this is typical of the Mira class) and more sinusoidal variation (also typical). Note that the Fourier methods appear to have estimated an incorrect period for this source. The true period appears to be around 161 days, half of the estimate.
\subsection{Classification Methodology}
There has been a lot of recent progress in the astronomy literature towards developing highly accurate classifiers for the time series generated by periodic variables \cite{debosscher2007automated,richards2011machine,dubath2011random}. The standard approach works as follows. A telescope observes a source $j$ at times $t_{1},\ldots,t_{l}$, recording flux measurements (how bright the object is) of $m_{1},\ldots,m_{l}$. Typically there are measurements of uncertainty on the flux measurements $e_{1},\ldots,e_{l}$. So each source $j$ is initially characterized by an $l$ tuples $D_j=\{(t_{i},m_{i},e_{i})\}_{i=1}^{l}$. Note that $l$, the number of times the source is observed, is different for each $D_j$. Also the time sampling of the flux measurements is irregular and different for each source. Associated with each source is a classification, such as Mira, Classical Cepheid, RR Lyrae, etc.

In order to construct a classifier, features are \textit{extracted} from $D_j$ (i.e. take functions of $D_j$) that will separate sources into different classes. Features vary from study to study, but typical ones include period (inverse of strongest Fourier frequency), amplitude, skew, and estimates of derivatives.

@@

\subsection{Uncertain Features}
\begin{figure}[h]
  \begin{center}
    \begin{includegraphics}[scale=.5]{period_amplitude.pdf}
      \caption{Period-Amplitude relationship for two classes: \textit{Classical Cepheid} stars and \textit{Population II Cepheid} stars. The green box represents (roughly) the physically possible range of periods and amplitudes. Clearly some of these features have been estimated incorrectly.\label{fig:period_amplitude}}
    \end{includegraphics}
  \end{center}
\end{figure}
A major problem with the standard approach is the high levels and heteroskedastic nature of the uncertainty in the features due to having poorly sampled time series ($l$ is small), high error in the flux measurements ($e_{i}$ are systematically large), or irregular nature of sampling ($t_i$'s are highly concentrated, giving a poor representation of variation). This leads to situations where features meant to represent physical quantities of the time series, such as amplitude or period, are incorrect. As an example in Figure \ref{fig:period_amplitude} we plot the period and amplitude features for two classes of stars with a box around known physical limits for these features for these two classes. Some of the observations have values outside these physical limits, clearly indicating that the features are incorrectly estimated. This suggests that classifiers which incorporate feature uncertainty may be able to achieve improved performance.

@@

This model, but without the $\frac{\lambda}{2}\beta^T\beta$ regularization, was considered in El Ghaoui \cite{el2003robust}. It is easy to show that for this form of $\mathcal{X}(\rho)$ one can explicitly solve for the inner max in equation \eqref{eq:svm_minimax} and obtain
\begin{equation}
\label{eq:svmInit}
\min_{\beta}\sum_{i=1}^n(1-y_i(\beta_0+\beta^Tx_i)+\rho\sigma_i^T|\beta|)_++\frac{\lambda}{2} ||\beta||_2^2.
\end{equation}
where $\sigma_i$ is the $i^{th}$ row of $\Sigma$. In \cite{el2003robust}, El Ghaoui drew similarities between the $\rho\sigma_i^T|\beta|$ in this expression and standard $L_1$ regularization for SVMs. For example, if $\Sigma_{i,j} = 1 \ \forall i,j$ then then expression becomes $\rho||\beta||_1$. This interpretation of $\mathcal{X}(\rho)$ in part motivates the use of $\rho$ as a tuning parameter rather than a quantity that is fixed by the inherent uncertainty in the data.

@@

From now on, we can safely assume that $n_+>n_-$, in which case the initial state is $\beta = 0$, $\beta_0=1$, $\xi_i = 0$ for $i\in \mathcal I_+$ and $\xi_i = 2$ for $i\in \mathcal I_-$. Thus by optimality conditions, for any $i\in \mathcal I_-$, $\gamma_i =0$ and so $\alpha_i=1$. In addition, since $0=\sum_{i=1}^n \alpha_i y_i = \sum_{i\in \mathcal I_+} \alpha_i - \sum_{i\in \mathcal I_-} \alpha_i$, we have $\sum_{i\in \mathcal I_+} \alpha_i = \sum_{i\in \mathcal I_-} \alpha_i = n_-$. The following lemma determines the initial value for the dual valuables:
\begin{lamma} Let $\tilde{\beta} = \sum_{i=1}^n \alpha_i y_i x_i + \nu - \mu$, 
\[
\begin{array}{rl}
(\alpha^*,\mu^*,\nu^*) = & \arg\min ||\tilde{\beta}||_2^2, \\
& \text{subject to: } 0\leq \alpha_i\leq 1,  \forall i \in \mathcal I_+; \alpha_i = 1,  \forall i \in \mathcal I_-; \\
& \sum_{i \in \mathcal I_+}\alpha_i = n_-, \rho\Sigma\alpha \geq \mu+\nu, \mu\geq0, \nu\geq0. \\
\end{array}
\]
 and $c^* = \rho\Sigma\alpha^* - \mu^*-\nu^*$. Then for some $\lambda_0$ we have for all $\lambda>\lambda_0$, $(\alpha(\lambda),\mu(\lambda),\nu(\lambda),c(\lambda)) = (\alpha^*, \mu^*,\nu^*,c^*)$.
\end{lamma}  
\noindent\textbf{Proof.} The proof follows directly from Lemma 2 in \cite{hastie2004entire}. The key observation is that  $\sum_{i=1}^n \alpha_i = 2n_-$ and this sum will remain constant for a while as $\beta$ grows from zero. Then we can plug this sum into the dual problem \eqref{eq:dual} and solve it with additional constraints: $\alpha_i = 1,  \forall i \in \mathcal I_-$ and $\sum_{i \in \mathcal I_+}\alpha_i = n_-$. \\

@@

%Remarks: in \cite{hastie2004entire}, Hastie et.al. split the initialization step into two cases depending on whether $n_+=n_-$ or $n_+>n_-$. In their case $n_+=n_-$, the initial state can be obtained efficiently without solving the quadratic programming problem. This does not seem to be the case in our problem, as it also involves finding the initial values of the extra dual variables $\mu$ and $\nu$. Therefore solving a quadratic problem for starting up the path algorithm cannot be avoided in our case.\\

@@

%\subsection{Extension to Kernel} (You need to transform the uncertainty in the original data space into the kernel space)
%\note{in fitting the CEP data,  between 2.03 and 2.13 $\beta0$ value can be arbitrary}
\subsection{Matlab and R programs}
All the programs for this project are included in the {\tt zip} file submitted together with this hard copy. In particular, the Matlab package {\tt svmIntervalPath} is the implementation of our path algorithm. Also, Matlab and R code for testing and data analysis can be found in that file as well.

\section{Application to Variable Star Data Sets}
\label{sec:application}
We now return to the periodic variable classification problem and discuss construction of hyper-rectangles, the specific data sets we use, and results from the models.
\subsection{Constructing the Intervals from the Time Series}
\begin{figure*}[ht]
\begin{center}
{\small
\framebox[6in]{
\begin{minipage}[t]{5.8in}
% Insert here your text
\begin{center} \textbf{Constructing Hyper-rectangles from Time Series} \end{center}
\begin{enumerate}
\item Order the light curve measurements in time i.e. $\{(t_1,m_1,e_1),(t_2,m_2,e_2), \ldots , (t_l,m_l,e_l)\}$ where $t_i < t_j$ for $i < j$.
\item Slice each light curve into 5 contiguous sections, producing 5 time series. Slice 1 is $(t_i,m_i,e_i)$ for $0 \leq i \leq .5l$, slice 2 for $.125l \leq i \leq .625l$, slice 3 for $.25l \leq i \leq .75l$ ect.
%% \begin{enumerate}
%% \item $\{(t_1,m_1,e_1), \ldots , (t_{l/2},m_{l/2},e_{l/2})\}$
%% \item $\{(t_1,m_1,e_1),(t_2,m_2,e_2), \ldots , (t_l,m_l,e_l)\}$
%% \end{enumerate}
\item Derive features for each of these slices.
\item Put the interval minimum at the lowest value of the feature for the 5 slices. Put the interval max at the maximum feature value across the 5 slices.
\end{enumerate}
\end{minipage}
}
}
\end{center}
\caption{Description of interval construction algorithm.\label{fig:interval_construction}}
\label{lb}
\end{figure*}
In certain applications, methods for constructing hyper-rectangles (or hyper-spheres) in feature space may be fairly strait forward. For example in El Ghaoui \cite{el2003robust}, the authors discuss applications to micro-array data where several replicates of some experiment are available, and a hyper-rectangle can be chosen to enclose all replicates.

@@

\begin{figure}
\include{RR2_features}
\caption{RR features with non-zero coefficients. Some coefficient are reported here as 0 due to rounding. Choosing a slightly higher $\rho$ may have caused several of the features with very small coefficients to become exactly 0, leading to a very sparse solution.\label{fig:RRfeatures}}
\end{figure}

\begin{figure}
\include{CEP_features}
\caption{CEP features with non-zero coefficients. Some coefficients are reported as 0 here due to rounding.\label{fig:CEPfeatures}}
\end{figure}

@@

\subsection{Conclusions}
We have developed an algorithm and empirically studied an SVM interval model which incorporates feature uncertainty and produces sparse classifiers. The model improves on classification performance over ignoring feature uncertainty. We feel that the results warrant applying this method to larger and multi-class astronomy data sets. From a data perspective we faced several challenges, the greatest of which was determining intervals for features. This will be common in many high dimensional classification problems where there is uncertainty on features, but no clear way to construct intervals. Generalizing feature extraction methods to produce estimates of uncertainty will make application of interval SVM methods much easier.

A natural extension of the SVM interval concept is to incorporate the feature uncertainty of the unknown test data point into the classifier. Broadly, if a new unlabeled observation has a lot of uncertainty on feature $x$, then it does not make sense to use feature $x$ heavily in the classifier. However if feature $x$ has small error then a classifier should make as much use of $x$ as possible. Since different observations will have different amounts of noise in different features, it may make sense to develop interval SVM methods that are specifically tuned the the uncertainty of each unlabeled observation. We will work on developing this idea further in future work.
